Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach

### 2014-05-14 manuscripts/manuscript.tex
track the tex intermediary file

---
layout: 12pt,review
linenumbers: true
title: "Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust ecosystem management.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points.  Before a collapse, we do not know
          where the tipping points lie, if they exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  These two sources of
          uncertainty frustrate state-of-the-art parametric approaches
          to decision theory and optimal control.  Here we illustrate
          how a non-parametric approach using a Gaussian process prior
          provides a more flexible representation of this inherent
          uncertainty. We can then embed the Gaussian process in a
          stochastic dynamic programming framework in order to make
          robust management predictions under both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  In keeping
          with previous results, we find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The Gaussian process approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches. We illustrate this using examples
          of simulated single-species dynamics, where the standard model
          selection approach should be most effective, and find that it
          still fails to account for uncertainty appropriately and leads
          to population crashes, while management based on the Gaussian 
          process approach does not. We discuss opportunities to adapt
          this approach to multiple species models in further work.

keywords:
  - Bayesian
  - Structural Uncertainty
  - Nonparametric 
  - Optimal Control
  - Decision Theory
  - Gaussian Processes
  - Fisheries Management
---

<!-- Run R code for analysis, to be called by figures -->
```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

```

```{r libraries, cache=FALSE, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
library("nonparametricbayes") 
library("pdgControl")

library("ggplot2") 
library("reshape2")
library("plyr")
library("data.table")

library("R2jags")
library("emdbook") # for as.mcmc.bugs (?)
library("coda")  # for as.mcmc

library("modeest")
library("MASS")
library("pander")

opts_chunk$set(fig.width=4, fig.height=3, echo=FALSE)
theme_set(theme_bw(base_size=12))
toggle = "hide" # results argument
```

```{r posterior-mode}
library(modeest)
posterior.mode <- function(x) {
  mlv(x, method="shorth")$M
}
```
```{r stateeq}
f <- RickerAllee
p <- c(2, 8, 5)
K <- 10  # approx, a li'l' less
allee <- 5 # approx, a li'l' less
```
```{r sdp-pars, dependson="stateeq"}
sigma_g <- 0.05
sigma_m <- 0.0
z_g <- function() rlnorm(1, 0, sigma_g)
z_m <- function() 1
x_grid <- seq(0, 1.5 * K, length=50)
h_grid <- x_grid
profit <- function(x,h) pmin(x, h)
delta <- 0.01
OptTime <- 50  # stationarity with unstable models is tricky thing
reward <- 0
xT <- 0
Xo <-  allee+.5# observations start from
x0 <- K # simulation under policy starts from
Tobs <- 40
MaxT = 1000 # timeout for value iteration convergence
```
```{r obs, dependson="sdp-pars", fig.keep='none'}
  set.seed(1234)
  #harvest <- sort(rep(seq(0, .5, length=7), 5))
  x <- numeric(Tobs)
  x[1] <- Xo
  nz <- 1
  for(t in 1:(Tobs-1))
    x[t+1] = z_g() * f(x[t], h=0, p=p)
  obs <- data.frame(x = c(rep(0,nz), 
                          pmax(rep(0,Tobs-1), x[1:(Tobs-1)])), 
                    y = c(rep(0,nz), 
                          x[2:Tobs]))
raw_plot <- ggplot(data.frame(time = 1:Tobs, x=x), aes(time,x)) + geom_line()
```

```{r mle, dependson="obs"}
set.seed(12345)
estf <- function(p){ 
    mu <- f(obs$x,0,p)
    -sum(dlnorm(obs$y, log(mu), p[4]), log=TRUE)
}
par <- c(p[1]*rlnorm(1,0,.1), 
         p[2]*rlnorm(1,0,.1), 
         p[3]*rlnorm(1,0, .1), 
         sigma_g * rlnorm(1,0,.1))
o <- optim(par, estf, method="L", lower=c(1e-5,1e-5,1e-5,1e-5))
f_alt <- f
p_alt <- c(as.numeric(o$par[1]), as.numeric(o$par[2]), as.numeric(o$par[3]))
sigma_g_alt <- as.numeric(o$par[4])

est <- list(f = f_alt, p = p_alt, sigma_g = sigma_g_alt, mloglik=o$value)

```
```{r mle-output, dependson="mle", results=toggle}
true_means <- sapply(x_grid, f, 0, p)
est_means <- sapply(x_grid, est$f, 0, est$p)
```

```{r gp-priors}
s2.p <- c(5,5)  
d.p = c(10, 1/0.1)
```
```{r gp, dependson=c("gp-priors", "obs")}
gp <- gp_mcmc(obs$x, y=obs$y, n=1e5, s2.p = s2.p, d.p = d.p)
gp_dat <- gp_predict(gp, x_grid, burnin=1e4, thin=300)
```
```{r gp_traces_densities, dependson="gp"}
gp_assessment_plots <- summary_gp_mcmc(gp, burnin=1e4, thin=300)
```
```{r gp-output, dependson="gp", results=toggle}
tgp_dat <- 
    data.frame(  x = x_grid, 
                 y = gp_dat$E_Ef, 
                 ymin = gp_dat$E_Ef - 2 * sqrt(gp_dat$E_Vf), 
                 ymax = gp_dat$E_Ef + 2 * sqrt(gp_dat$E_Vf) )
```

```{r jags-setup}
y <- x 
N <- length(x);
jags.data <- list("N","y")
n.chains <- 6
n.iter <- 1e6
n.burnin <- floor(10000)
n.thin <- max(1, floor(n.chains * (n.iter - n.burnin)/1000))
n.update <- 10

```
```{r common-priors}
stdQ_prior_p <- c(1e-6, 100)
stdR_prior_p <- c(1e-6, .1)
stdQ_prior  <- function(x) dunif(x, stdQ_prior_p[1], stdQ_prior_p[2])
stdR_prior  <- function(x) dunif(x, stdR_prior_p[1], stdR_prior_p[2])

```
```{r allen-model}
K_prior_p <- c(0.01, 20.0)
r0_prior_p <- c(0.01, 6.0)
theta_prior_p <- c(0.01, 20.0)

bugs.model <- 
paste(sprintf(
"model{
  K     ~ dunif(%s, %s)
  r0    ~ dunif(%s, %s)
  theta ~ dunif(%s, %s)
  stdQ ~ dunif(%s, %s)", 
  K_prior_p[1], K_prior_p[2],
  r0_prior_p[1], r0_prior_p[2],
  theta_prior_p[1], theta_prior_p[2],
  stdQ_prior_p[1], stdQ_prior_p[2]),

  "
  iQ <- 1 / (stdQ * stdQ);
  y[1] ~ dunif(0, 10)
  for(t in 1:(N-1)){
    mu[t] <- log(y[t]) + r0 * (1 - y[t]/K)* (y[t] - theta) / K 
    y[t+1] ~ dlnorm(mu[t], iQ) 
  }
}")
writeLines(bugs.model, "allen_process.bugs")

```
```{r allen-priors, dependson="common-priors"}
K_prior     <- function(x) dunif(x, K_prior_p[1], K_prior_p[2])
r0_prior <- function(x) dunif(x, r0_prior_p[1], r0_prior_p[2])
theta_prior <- function(x) dunif(x, theta_prior_p[1], theta_prior_p[2])
par_priors  <- list(K = K_prior, deviance = function(x) 0 * x, 
                    r0 = r0_prior, theta = theta_prior,
                    stdQ = stdQ_prior)

```

```{r allen_priors_xtable, include=TRUE, echo=FALSE}
allen_priors_xtable <- data.frame(parameter = c("$r$", "$K$", "$X_C$", "$\\sigma$"),
           "lower bound" = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           "upper bound" = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2]))
```

```{r allen-mcmc, dependson=c("allen-model", "allen-pars", "jags-setup"), results=toggle}
jags.params=c("K","r0","theta","stdQ") # be sensible about the order here
jags.inits <- function(){
  list("K"= 10 * rlnorm(1,0, 0.1),
       "r0"= 1 * rlnorm(1,0, 0.1) ,
       "theta"=   5 * rlnorm(1,0, 0.1) , 
       "stdQ"= abs( 0.1 * rlnorm(1,0, 0.1)),
       .RNG.name="base::Wichmann-Hill", .RNG.seed=123)
}

set.seed(1234)
# parallel refuses to take variables as arguments (e.g. n.iter = 1e5 works, but n.iter = n doesn't)
allen_jags <- do.call(jags, list(data=jags.data, inits=jags.inits, 
                                      jags.params, n.chains=n.chains, 
                                      n.iter=n.iter, n.thin=n.thin, 
                                      n.burnin=n.burnin, 
                                      model.file="allen_process.bugs"))

# Run again iteratively if we haven't met the Gelman-Rubin convergence criterion
recompile(allen_jags) # required for parallel
allen_jags <- do.call(autojags, 
											list(object=allen_jags, n.update=n.update, 
                           n.iter=n.iter, n.thin = n.thin))

```
```{r allen-traces, dependson="allen-mcmc"}
tmp <- lapply(as.mcmc(allen_jags), as.matrix) # strip classes the hard way...
allen_posteriors <- melt(tmp, id = colnames(tmp[[1]])) 
names(allen_posteriors) = c("index", "variable", "value", "chain")
plot_allen_traces <- ggplot(allen_posteriors) + geom_line(aes(index, value)) + 
  facet_wrap(~ variable, scale="free", ncol=1)

```
```{r allen-posteriors, dependson=c("allen-traces", "allen-priors")}
allen_priors <- ddply(allen_posteriors, "variable", function(dd){
    grid <- seq(min(dd$value), max(dd$value), length = 100) 
    data.frame(value = grid, density = par_priors[[dd$variable[1]]](grid))
})
plot_allen_posteriors <- ggplot(allen_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
#  geom_line(data=allen_priors, aes(x=value, y=density), col="red") +  
  facet_wrap(~ variable, scale="free", ncol=3)

```
```{r allen-output, dependson=c("posterior-mode", "allen-traces"), results=toggle}
A <- allen_posteriors
A$index <- A$index + A$chain * max(A$index) # Combine samples across chains by renumbering index 
pardist <- acast(A, index ~ variable)
bayes_coef <- apply(pardist,2, posterior.mode) 
bayes_pars <- unname(c(bayes_coef["r0"], bayes_coef["K"], bayes_coef["theta"])) # parameters formatted for f
allen_f <- function(x,h,p) unname(RickerAllee(x,h, unname(p[c("r0", "K", "theta")])))
allen_means <- sapply(x_grid, f, 0, bayes_pars)
bayes_pars
head(pardist)

```

```{r ricker-model}
K_prior_p <- c(0.01, 40.0)
r0_prior_p <- c(0.01, 20.0)
bugs.model <- 
paste(sprintf(
"model{
  K    ~ dunif(%s, %s)
  r0    ~ dunif(%s, %s)
  stdQ ~ dunif(%s, %s)", 
  K_prior_p[1], K_prior_p[2],
  r0_prior_p[1], r0_prior_p[2],
  stdQ_prior_p[1], stdQ_prior_p[2]),
  "
  iQ <- 1 / (stdQ * stdQ);
  y[1] ~ dunif(0, 10)
  for(t in 1:(N-1)){
    mu[t] <- log(y[t]) + r0 * (1 - y[t]/K) 
    y[t+1] ~ dlnorm(mu[t], iQ) 
  }
}")
writeLines(bugs.model, "ricker_process.bugs")

```
```{r ricker-priors, dependson="common-priors"}
K_prior     <- function(x) dunif(x, K_prior_p[1], K_prior_p[2])
r0_prior <- function(x) dunif(x, r0_prior_p[1], r0_prior_p[2])
par_priors <- list(K = K_prior, deviance = function(x) 0 * x, 
                   r0 = r0_prior, stdQ = stdQ_prior)

```

```{r ricker_priors_xtable, include=TRUE, echo=FALSE, cache=FALSE}
ricker_priors_xtable <- data.frame(
  parameter = c("$r$", "$K$", "$\\sigma$"),
  "lower bound" = c(r0_prior_p[1], K_prior_p[1], stdQ_prior_p[1]),
  "upper bound" = c(r0_prior_p[2], K_prior_p[2], stdQ_prior_p[2]))
```

```{r ricker-mcmc, dependson="ricker-model", results=toggle}
jags.params=c("K","r0", "stdQ")
jags.inits <- function(){
  list("K"= 10 * rlnorm(1,0,.5),
       "r0"= rlnorm(1,0,.5),
       "stdQ"=sqrt(0.05) * rlnorm(1,0,.5),
       .RNG.name="base::Wichmann-Hill", .RNG.seed=123)
}
set.seed(12345) 
ricker_jags <- do.call(jags, 
                       list(data=jags.data, inits=jags.inits, 
                            jags.params, n.chains=n.chains, 
                            n.iter=n.iter, n.thin=n.thin, n.burnin=n.burnin,
                            model.file="ricker_process.bugs"))
recompile(ricker_jags)
ricker_jags <- do.call(autojags, 
                       list(object=ricker_jags, n.update=n.update, 
														n.iter=n.iter, n.thin = n.thin, 
														progress.bar="none"))

```
```{r ricker-traces, dependson="ricker-mcmc"}
tmp <- lapply(as.mcmc(ricker_jags), as.matrix) # strip classes the hard way...
ricker_posteriors <- melt(tmp, id = colnames(tmp[[1]])) 
names(ricker_posteriors) = c("index", "variable", "value", "chain")
plot_ricker_traces <- ggplot(ricker_posteriors) + geom_line(aes(index, value)) + 
  facet_wrap(~ variable, scale="free", ncol=1)
```
```{r ricker-posteriors, dependson=c("ricker-traces", "ricker-priors")}
ricker_priors <- ddply(ricker_posteriors, "variable", function(dd){
    grid <- seq(min(dd$value), max(dd$value), length = 100) 
    data.frame(value = grid, density = par_priors[[dd$variable[1]]](grid))
})
# plot posterior distributions
plot_ricker_posteriors <- ggplot(ricker_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
#  geom_line(data=ricker_priors, aes(x=value, y=density), col="red") +  # don't plot priors 
  facet_wrap(~ variable, scale="free", ncol=2)
```
```{r ricker-output, dependson=c("posterior-mode", "ricker-traces"), results=toggle}
A <- ricker_posteriors
A$index <- A$index + A$chain * max(A$index) # Combine samples across chains by renumbering index 
ricker_pardist <- acast(A, index ~ variable)
bayes_coef <- apply(ricker_pardist,2, posterior.mode) 
ricker_bayes_pars <- unname(c(bayes_coef["r0"], bayes_coef["K"]))
ricker_f <- function(x,h,p){
  sapply(x, function(x){ 
    x <- pmax(0, x-h) 
    pmax(0, x * exp(p["r0"] * (1 - x / p["K"] )) )
  })
}
ricker_means <- sapply(x_grid, Ricker, 0, ricker_bayes_pars[c(1,2)])
head(ricker_pardist)
ricker_bayes_pars

```

```{r myers-model}
r0_prior_p <- c(.0001, 10.0)
theta_prior_p <- c(.0001, 10.0)
K_prior_p <- c(.0001, 40.0)
bugs.model <- 
paste(sprintf(
"model{
  r0    ~ dunif(%s, %s)
  theta    ~ dunif(%s, %s)
  K    ~ dunif(%s, %s)
  stdQ ~ dunif(%s, %s)", 
  r0_prior_p[1], r0_prior_p[2],
  theta_prior_p[1], theta_prior_p[2],
  K_prior_p[1], K_prior_p[2],
  stdQ_prior_p[1], stdQ_prior_p[2]),

  "
  iQ <- 1 / (stdQ * stdQ);

  y[1] ~ dunif(0, 10)
  for(t in 1:(N-1)){
    mu[t] <- log(r0)  + theta * log(y[t]) - log(1 + pow(abs(y[t]), theta) / K)
    y[t+1] ~ dlnorm(mu[t], iQ) 
  }
}")
writeLines(bugs.model, "myers_process.bugs")

```
```{r myers-priors}
K_prior     <- function(x) dunif(x, K_prior_p[1], K_prior_p[2])
r_prior     <- function(x) dunif(x, r0_prior_p[1], r0_prior_p[2])
theta_prior <- function(x) dunif(x, theta_prior_p[1], theta_prior_p[2])
par_priors <- list( deviance = function(x) 0 * x, K = K_prior,
                    r0 = r_prior, theta = theta_prior, 
                    stdQ = stdQ_prior)

```

```{r myers_prior_xtable, include=TRUE, echo=FALSE} 
myers_priors_xtable <- data.frame(parameter = c("$r$", "$K$", "$\\theta$", "$\\sigma$"),
           "lower bound" = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           "upper bound" = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2]))
```

```{r myers-mcmc, dependson="myers-model", results=toggle}
jags.params=c("r0", "theta", "K", "stdQ")
jags.inits <- function(){
  list("r0"= 1 * rlnorm(1,0,.1), 
       "K"=    10 * rlnorm(1,0,.1),
       "theta" = 1 * rlnorm(1,0,.1),  
       "stdQ"= sqrt(0.2) * rlnorm(1,0,.1),
       .RNG.name="base::Wichmann-Hill", .RNG.seed=123)
}
set.seed(12345)
myers_jags <- do.call(jags, 
                      list(data=jags.data, inits=jags.inits, 
													 jags.params, n.chains=n.chains, 
													 n.iter=n.iter, n.thin=n.thin,
                           n.burnin=n.burnin, 
                           model.file="myers_process.bugs"))
recompile(myers_jags)
myers_jags <- do.call(autojags, 
                      list(myers_jags, n.update=n.update, 
                           n.iter=n.iter, n.thin = n.thin, 
                           progress.bar="none"))

```
```{r myers-traces, dependson="myers-mcmc"}
tmp <- lapply(as.mcmc(myers_jags), as.matrix) # strip classes
myers_posteriors <- melt(tmp, id = colnames(tmp[[1]])) 
names(myers_posteriors) = c("index", "variable", "value", "chain")
plot_myers_traces <- ggplot(myers_posteriors) + 
  geom_line(aes(index, value)) + # priors, need to fix order though
  facet_wrap(~ variable, scale="free", ncol=1)

```
```{r myers-posteriors, dependson="myers-traces"}
par_prior_curves <- ddply(myers_posteriors, "variable", function(dd){
    grid <- seq(min(dd$value), max(dd$value), length = 100) 
    data.frame(value = grid, density = par_priors[[dd$variable[1]]](grid))
})
plot_myers_posteriors <- ggplot(myers_posteriors, aes(value)) + 
  stat_density(geom="path", position="identity", alpha=0.7) +
#  geom_line(data=par_prior_curves, aes(x=value, y=density), col="red") +  # Whoops, these are misaligned. see table instead 
  facet_wrap(~ variable, scale="free", ncol=3)

```
```{r myers-output, dependson=c("posterior-mode", "myers-traces"), results=toggle}
A <- myers_posteriors
A$index <- A$index + A$chain * max(A$index) # Combine samples across chains by renumbering index 
myers_pardist <- acast(A, index ~ variable)
bayes_coef <- apply(myers_pardist,2, posterior.mode) # much better estimates
myers_bayes_pars <- unname(c(bayes_coef["r0"], bayes_coef["theta"], bayes_coef["K"]))
myers_means <- sapply(x_grid, Myer_harvest, 0, myers_bayes_pars)
myers_f <- function(x,h,p) Myer_harvest(x, h, p[c("r0", "theta", "K")])
head(myers_pardist)
myers_bayes_pars

```

```{r assemble-models, dependson=c("myers-output", "ricker-output", "allen-output", "gp-output", "mle-output")}
models <- data.frame(x=x_grid, 
										 GP=tgp_dat$y, 
										 True=true_means, 
                     MLE=est_means, 
										 Ricker=ricker_means, 
                     Allen = allen_means,
                     Myers = myers_means)
models <- melt(models, id="x")
# some labels
names(models) <- c("x", "method", "value")
# labels for the colorkey too
model_names = c("GP", "True", "MLE", "Ricker", "Allen", "Myers")
colorkey=cbPalette
names(colorkey) = model_names 
step_ahead <- function(x, f, p){
  h = 0
  x_predict <- sapply(x, f, h, p)
  n <- length(x_predict) - 1
  y <- c(x[1], x_predict[1:n])
  y
}
```
```{r gp-opt, dependson="gp-output"}
matrices_gp <- gp_transition_matrix(gp_dat$Ef_posterior, gp_dat$Vf_posterior, x_grid, h_grid) 
opt_gp <- value_iteration(matrices_gp, x_grid, h_grid, MaxT, xT, profit, delta, reward)
```
```{r mle-opt, dependson="mle-output"}
matrices_true <- f_transition_matrix(f, p, x_grid, h_grid, sigma_g)
opt_true <- value_iteration(matrices_true, x_grid, h_grid, OptTime=MaxT, xT, profit, delta=delta)
matrices_estimated <- f_transition_matrix(est$f, est$p, x_grid, h_grid, est$sigma_g)
opt_estimated <- value_iteration(matrices_estimated, x_grid, h_grid, OptTime=MaxT, xT, profit, delta=delta)

```
```{r allen-opt, dependson="allen-output"}
matrices_allen <- parameter_uncertainty_SDP(allen_f, x_grid, h_grid, pardist, 4)
opt_allen <- value_iteration(matrices_allen, x_grid, h_grid, OptTime=MaxT, xT, profit, delta=delta)
```
```{r ricker-opt, dependson="ricker-output"}
matrices_ricker <- parameter_uncertainty_SDP(ricker_f, x_grid, h_grid, as.matrix(ricker_pardist), 3)
opt_ricker <- value_iteration(matrices_ricker, x_grid, h_grid, OptTime=MaxT, xT, profit, delta=delta)
```
```{r myers-opt, dependson="myers-output"}
matrices_myers <- parameter_uncertainty_SDP(myers_f, x_grid, h_grid, as.matrix(myers_pardist), 4)
myers_alt <- value_iteration(matrices_myers, x_grid, h_grid, OptTime=MaxT, xT, profit, delta=delta)

```
```{r assemble-opt, dependson=c("gp-opt", "mle-opt", "allen-opt", "ricker-opt", "myers-opt")}
OPT = data.frame(GP = opt_gp$D, True = opt_true$D, MLE = opt_estimated$D, Ricker = opt_ricker$D, Allen = opt_allen$D, Myers = myers_alt$D)
colorkey=cbPalette
names(colorkey) = names(OPT) 
```
```{r sims, dependson="assemble-opt"}
sims <- lapply(OPT, function(D){
  set.seed(1)
  lapply(1:100, function(i) 
    ForwardSimulate(f, p, x_grid, h_grid, x0, D, z_g, profit=profit, OptTime=OptTime)
  )
})
# turn the list into a data.frame
keep.as.columns <- names(sims[[1]][[1]])
dat <- melt(sims, id=keep.as.columns)
sims_data <- data.table(dat)
setnames(sims_data, c("L1", "L2"), c("method", "reps")) 
# Legend in original ordering please, not alphabetical: 
sims_data$method = factor(sims_data$method, ordered=TRUE, levels=names(OPT))
```

```{r profits, dependson="sims", results=toggle}
Profit <- sims_data[, sum(profit), by=c("reps", "method")]
tmp <- dcast(Profit, reps ~ method)
#tmp$Allen <- tmp[,"Allen"] + rnorm(dim(tmp)[1], 0, 1) # jitter for plotting
tmp <- tmp / tmp[,"True"]
tmp <- melt(tmp[2:dim(tmp)[2]])
actual_over_optimal <-subset(tmp, variable != "True")
```

>>>>

\documentclass[author-year, 12pt,review]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
  \linenumbers % turns line numbering on 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}

  \title{Avoiding tipping points in ecosystem based fisheries management: a
non-parametric Bayesian approach}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig@gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[cstar]{Marc Mangel}

    \author[noaa]{Stephan Munch}

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric
Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}

  \begin{abstract}
  Model uncertainty and limited data coverage are fundamental challenges
  to robust ecosystem management. These challenges are acutely highlighted
  by concerns that many ecological systems may contain tipping points.
  Before a collapse, we do not know where the tipping points lie, if they
  exist at all. Hence, we know neither a complete model of the system
  dynamics nor do we have access to data in some large region of
  state-space where such a tipping point might exist. These two sources of
  uncertainty frustrate state-of-the-art parametric approaches to decision
  theory and optimal control. Here we illustrate how a non-parametric
  approach using a Gaussian process prior provides a more flexible
  representation of this inherent uncertainty. We can then embed the
  Gaussian process in a stochastic dynamic programming framework in order
  to make robust management predictions under both model uncertainty and
  limited data. We conduct a set of simulations to evaluate this approach
  as compared with the standard approach of using model selection to
  choose from a set of candidate models. In keeping with previous results,
  we find that model selection erroneously favors models without tipping
  points -- leading to harvest policies that guarantee extinction. The
  Gaussian process approach performs nearly as well as the true model and
  significantly outperforms standard appraoches. We illustrate this using
  examples of simulated single-species dynamics, where the standard model
  selection approach should be most effective, and find that it still
  fails to account for uncertainty appropriately and leads to population
  crashes, while management based on the Gaussian process approach does
  not. We discuss opportunities to adapt this approach to multiple species
  models in further work.
  \end{abstract}
   \begin{keyword} Bayesian \sep Structural Uncertainty \sep Nonparametric \sep Optimal Control \sep Decision Theory \sep Gaussian Processes \sep Fisheries Management \sep \end{keyword}
 \end{frontmatter}

\section{Introduction}\label{introduction}

Decision making under uncertainty is a ubiquitous challenge of natural
resource management and conservation. Ecological dynamics are frequently
complex and difficult to measure, making uncertainty in our
understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the best
sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified (Fischer \emph{et al.} 2009).
Concerns about the potential sudden collapse of fisheries and other
ecosystems (Worm \emph{et al.} 2006; Scheffer \emph{et al.} 2009) has
increasingly emphasized the difficulties of formulating even
qualitatively correct models of the underlying processes.

We develop a novel approach to address these concerns in the context of
fisheries; though the underlying challenges and methods are germane to
many other conservation and resource management problems. Economic value
and ecological concern have made marine fisheries the crucible for much
of the founding work (Gordon 1954; May \emph{et al.} 1979; Reed 1979;
Ludwig \& Walters 1982) on management under uncertainty. Global declines
(Worm \emph{et al.} 2006) and the controversy surrounding their
interpretation (Hilborn 2007; Worm \emph{et al.} 2009) have made
understanding these challenges all the more pressing.

Uncertainty enters the decision-making process at many levels: intrinsic
stochasticity in biological processes, measurements, and implementation
of policy (\emph{e.g.} Reed 1979; Clark \& Kirkwood 1986; Roughgarden \&
Smith 1996; Sethi \emph{et al.} 2005), parametric uncertainty
(\emph{e.g.} Ludwig \& Walters 1982; Hilborn \& Mangel 1997; McAllister
1998; Schapaugh \& Tyre 2013), and model or structural uncertainty
(\emph{e.g.} Williams 2001; Cressie \emph{et al.} 2009; Athanassoglou \&
Xepapadeas 2012). Of these, structural uncertainty is generally the
hardest to quantify. Typical approaches assume a weak notion of model
uncertainty in which the correct model (or reasonable approximation) of
the dynamics must be identified from among a handful of alternative
models. Here we consider an approach that addresses uncertainty at each
of these levels without assuming the dynamics follow a particular
(i.e.~parametric) structure.

Often we do not understand the system dynamics well enough to know if a
model provides a good approximation over the relevant range of states
and timescales (criteria that we loosely refer to as defining the
``right'' or ``true'' model.)\\Typical solutions involve either model
choice, model averaging, or introducing yet greater model complexity of
which others may be special cases (model averaging being one such way to
construct such a model) (Williams 2001; Cressie \emph{et al.} 2009;
Athanassoglou \& Xepapadeas 2012). Even setting aside other
computational and statistical concerns (e.g. (Cressie \emph{et al.}
2009)), these approaches do not address our second concern -
representing uncertainty outside the observed data range.

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated.
This extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b)
decision-theoretic algorithms rely not only on reasonable estimates of
the expected outcomes, but depend on the weights given to all possible
outcomes (\emph{e.g.} Weitzman 2013).

This difficult position of having neither the true model nor data that
covers the full range of possible states is unfortunately the rule more
than the exception. The potential concern of tipping points in
ecological dynamics (Scheffer \emph{et al.} 2001; Polasky \emph{et al.}
2011) reflects these concerns -- as either knowledge of the true model
or more complete sampling of the state space would make it easy to
identify if a tipping point existed. If we do not know but cannot rule
out such a possibility, then we face decision-making under this dual
challege of model uncertainty and incomplete data coverage.

These dual concerns pose a substantial challenge to existing
decision-theoretic approaches (Brozović \& Schlenker 2011). Because
intervention is often too late after a tipping point has been crossed
(but see Hughes et al. (2013)), management is most often concerned with
avoiding potentially catastrophic tipping points before any data is
available at or following a transition that would more clearly reveal
these regime shift dynamics (e.g. Bestelmeyer \emph{et al.} 2012).

Here we illustrate how a stochastic dynamic programming (SDP) algorithm
(Mangel \& Clark 1988; Marescot \emph{et al.} 2013) can be driven by the
predictions from a Bayesian non-parametric (BNP) approach (Munch
\emph{et al.} 2005a). This provides two distinct advantages compared
with contemporary approaches. First, using a BNP sidesteps the need for
an accurate model-based description of the system dynamics. Second, the
BNP can better reflect uncertainty that arises when extrapolating a
model outside of the data on which it was fit. We illustrate that when
the correct model is not known, this latter feature is crucial to
providing a robust decision-theoretic approach in face of substantial
structural uncertainty.
